---
title: "pumps"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{pumps}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
CAN_RUN_JAGS <- requireNamespace("rjags", quietly = TRUE) && nzchar(Sys.which("jags"))
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 7, fig.height = 4.8,
  message = FALSE, warning = FALSE
)
suppressPackageStartupMessages({
  library(dplyr); library(tidyr); library(ggplot2); library(coda)
  library(rjags); library(knitr)
  library(dagnabit.docs)
})
set.seed(123)
```

---
title: "Pumps — Original BUGS → Official Diagram → Package Recreation"
author: "Naz Yucel"
date: "`r format(Sys.Date())`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Pumps — Original BUGS → Official Diagram → Package Recreation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

### Overview
The Pumps example models failures in ten power-plant pumps using a hierarchical Gamma–Poisson setup that naturally accounts for differing exposure times. Each pump’s observed failures (x_i) follow a Poisson distribution with mean (lambda_i) = (theta_i) * (t_i), where (theta_i) is the failure rate for pump i and (t_i) is the length of operation time of the pump. To borrow strength across pumps, the rates share a common prior (theta_i) ~ Gamma(alpha, beta), with weakly informative hyperpriors such as (alpha) ~ Exponential(1.0) and (beta) ~ Gamma(0.1, 1.0). Fitted with JAGS, the model yields posterior estimates for the population hyperparameters (alpha), (beta) and each pump's rate (theta_i), delivering partial pooling that stabilizes noisy per-pump estimates when data are sparse. Typical outputs include a summary table of posteriors and a simple fit check plotting observed counts (x_i) against posterior mean (lambda_i) (points near the 45° line indicate a good match).

### A) Original BUGS model

> The Pumps example (conjugate Gamma–Poisson hierarchical model) as presented in the WinBUGS help examples.

```{r pumps-bugs, results='asis'}
paths <- c(
  system.file("extdata/winbugs/pumps_model.bug", package = "dagnabit.docs"),
  "inst/extdata/winbugs/pumps_model.bug",
  "../inst/extdata/winbugs/pumps_model.bug",
  "../../inst/extdata/winbugs/pumps_model.bug"
)
paths <- paths[nzchar(paths)]
p <- paths[which(file.exists(paths))[1]]
if (!length(p)) {
  cat("Missing file: Please add inst/extdata/winbugs/pumps_model.bug to the package and re-knit.\n")
} else {
  cat(paste(readLines(p, warn = FALSE), collapse = "\n"))
}
```

### B) Official WinBUGS diagram

```{r pumps-diagram, echo=FALSE, fig.cap="Official WinBUGS 'Pumps' graphical model"}
paths <- c(
system.file("extdata/winbugs/pumps_diagram.png", package = "dagnabit.docs"),
"inst/extdata/winbugs/pumps_diagram.png",
"../inst/extdata/winbugs/pumps_diagram.png",
"../../inst/extdata/winbugs/pumps_diagram.png"
)
paths <- paths[nzchar(paths)]
img <- paths[which(file.exists(paths))[1]]
if (!length(img)) {
cat("Diagram not found. Add `inst/extdata/winbugs/pumps_diagram.png` and re-knit.\n")
} else {
knitr::include_graphics(img)
}
```

## C) Package recreation (fit + outputs)

We fit the Gamma–Poisson hierarchical model with JAGS:
xi ~ Poisson(theta_i*t_i), i = 1,...,10
theta_i ~ Gamma(alpha, beta), i = 1,...,10
alpha ~ Exponential(1.0)
beta ~ Gamma (0.1, 1.0)

Operation times (thousands of hours) and failure counts from WinBUGS help.

## Data
```{r pumps-data}

t <- c(94.3, 15.7, 62.9, 126.0, 5.24, 31.4, 1.05, 1.05, 2.1, 10.5)
x <- c(   5,   1,    5,   14,   3,   19,   1,   1,   4,   22)
N <- length(x)

pumps_df <- tibble::tibble(pump = factor(seq_len(N)), t, x)
knitr::kable(pumps_df, caption = "Pumps data (operation time t and failures x).")
```

### JAGS model 

```{r pumps-model}
pumps_model <- "
model {
for (i in 1:N) {
x[i] ~ dpois(lambda[i])
lambda[i] <- theta[i] * t[i]
theta[i] ~ dgamma(alpha, beta)
}

# hyperpriors

alpha ~ dexp(1.0)
beta  ~ dgamma(0.1, 1.0)
}
"

jags_data <- list(x = x, t = t, N = N)

# mild, dispersed inits

make_inits <- function() list(
theta = pmax(x / pmax(t, 1e-6), 1e-3),  # crude rates
alpha = 1,
beta  = 1
)

params <- c("alpha","beta","theta")
```

### Fit & sample

```{r pumps-fit, eval=CAN_RUN_JAGS, results='hide', message=TRUE}
mod <- rjags::jags.model(
textConnection(pumps_model),
data  = jags_data,
inits = list(make_inits(), make_inits(), make_inits()),
n.chains = 3,
n.adapt  = 1000
)

update(mod, 2000)  # burn-in
samp <- rjags::coda.samples(
model = mod,
variable.names = params,
n.iter = 10000,
thin = 5
)
```

```{r pumps-summaries, eval=exists("samp")}
s <- summary(samp)
tab <- cbind(
  s$statistics[, c("Mean","SD")],
  s$quantiles[, c("2.5%","50%","97.5%")]
)
wanted <- c("alpha","beta", paste0("theta[", seq_len(N), "]"))
knitr::kable(round(tab[wanted, , drop=FALSE], 4),
             caption = "Posterior summaries for alpha, beta and theta[1:N].")
```

### Diagnostics

```{r pumps-diagnostics, eval=exists("samp")}
coda::gelman.diag(samp[, c("alpha","beta")])
coda::autocorr.plot(samp[, c("alpha","beta")])
```

### Fitted lines vs data

```{r pumps-fitted, eval=exists("samp")}
m <- do.call(rbind, samp)

# posterior means of theta[i]

theta_cols <- grepl("^theta\\[[0-9]+\\]$", colnames(m))
theta_mean <- colMeans(m[, theta_cols, drop = FALSE])

plot_df <- tibble::tibble(
pump = factor(seq_len(N)),
t = t,
x = x,
theta_hat = as.numeric(theta_mean),
lambda_hat = theta_hat * t
)

ggplot2::ggplot(plot_df, ggplot2::aes(x = lambda_hat, y = x, label = pump)) +
ggplot2::geom_abline(slope = 1, intercept = 0, linetype = 2) +
ggplot2::geom_point(size = 2) +
ggplot2::geom_text(nudge_y = 0.4, size = 3) +
ggplot2::labs(x = expression("Posterior mean " * lambda[i] == theta[i] %.% t[i]),
              y = expression("Observed failures " * x[i]))
```



